The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2009-07-01 script/en/chapters/primes.tex
chaper on primes: port Pari/GP code to Sage

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{Appendix_A}{}
\section*{Appendix A: Examples using Pari-GP}
\addcontentsline{toc}{section}{Appendix A: Examples using Pari-GP}
\index{Pari-GP}

This appendix contains the source code for Pari-GP to compute e.g. the
quadratic sample sequences for primes.

\hypertarget{PrimArith1}{}
\subsection*{Validation of the results of a quadratic ``prime number formula''}

Sequences like these in the function value table on page \pageref{SrcPrimArith1}
can be computed in Pari-GP with the following commands:

\vspace{-2ex}
\begin{verbatim}
print("N -- N^2 - 9*N + 61");
for(N=0,50, X=N^2 - 9*N + 61;
  if (isprime(X)==1,
    print(N, " -- ", X, "  is prime"),
    print(N, " -- ", X, "  is NOT prime")
     );
   );
\end{verbatim}

The output looks like this:
\vspace{-2ex}
\begin{verbatim}
N -- N^2 - 9*N + 61
0 -- 61  is prime
1 -- 53  is prime
2 -- 47  is prime
3 -- 43  is prime
4 -- 41  is prime
5 -- 41  is prime
6 -- 43  is prime
7 -- 47  is prime
8 -- 53  is prime
9 -- 61  is prime
10 -- 71  is prime
11 -- 83  is prime
12 -- 97  is prime
13 -- 113  is prime
14 -- 131  is prime
15 -- 151  is prime
16 -- 173  is prime
17 -- 197  is prime
18 -- 223  is prime
19 -- 251  is prime
20 -- 281  is prime
21 -- 313  is prime
22 -- 347  is prime
23 -- 383  is prime
24 -- 421  is prime
25 -- 461  is prime
26 -- 503  is prime
27 -- 547  is prime
28 -- 593  is prime
29 -- 641  is prime
30 -- 691  is prime
31 -- 743  is prime
32 -- 797  is prime
33 -- 853  is prime
34 -- 911  is prime
35 -- 971  is prime
36 -- 1033  is prime
37 -- 1097  is prime
38 -- 1163  is prime
39 -- 1231  is prime
40 -- 1301  is prime
41 -- 1373  is prime
42 -- 1447  is prime
43 -- 1523  is prime
44 -- 1601  is prime
45 -- 1681  is NOT prime
46 -- 1763  is NOT prime
47 -- 1847  is prime
48 -- 1933  is prime
49 -- 2021  is NOT prime
50 -- 2111  is prime
\end{verbatim}

\hypertarget{PrimArith2}{}
\subsection*{Percentage of primes, delivered by a quadratic ``prime number formula''}

Pari-GP command:
\vspace{-2ex}
\begin{verbatim}
A=0;Beg=0;End=1000;
for(N=Beg,End, X=N^2-9*N+61;
  if(isprime(X)==1, 
  print(N, " -- ", X, "  is prime");A++,
  print(N, " -- ", X, "  is NOT prime")
    ) );
print("Number of primes: ", A, ";  Percentage of primes: " (A*100)/((End-Beg)+1)*1.);
\end{verbatim}
The output looks like this:
\vspace{-2ex}
\begin{verbatim}
...
1000 -- 991061  is NOT prime
Number of primes: 584;  Percentage of primes: 58.34165834165834165834165834
\end{verbatim}

>>>>

% --------------------------------------------------------------------------
\newpage
\hypertarget{Appendix_A}{}
\section*{Appendix A: Examples using Sage}
\addcontentsline{toc}{section}{Appendix A: Examples using Sage}
\index{Sage}

This appendix contains Sage source code to verify primality of
integers generated by the function $f(n) = n^2 - 9n + 61$. The code
defines a function called \verb!quadratic_prime_formula()! that takes
three arguments:
\begin{itemize}
\item \verb!start! --- An integer which is the lower bound for
  integers in the sequence $\texttt{start}, \texttt{start} + 1,
  \texttt{start} + 2, \dots, \texttt{end} - 1, \texttt{end}$.

\item \verb!end! --- An integer which is the upper bound for the
  integers in the sequence $\texttt{start}, \texttt{start} + 1,
  \texttt{start} + 2, \dots, \texttt{end} - 1, \texttt{end}$.

\item \verb!verbose! --- (default: \verb!True!) a flag to signify
  whether to print a message indicating the primality of an integer
  generated by $f(n)$.
\end{itemize}
The code can be modified to verify the primality of integers generated
by other functions that take integer values.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def quadratic_prime_formula(start, end, verbose=True):
....:     print "N -- N^2 - 9*N + 61"
....:     P = 0  # the number of primes between start and end
....:     for n in xrange(start, end + 1):
....:         X = n^2 - 9*n + 61
....:         if is_prime(X):
....:             P += 1
....:             if verbose:
....:                 print str(n) + " -- " + str(X) + " is prime"
....:         else:
....:             if verbose:
....:                 print str(n) + " -- " + str(X) + " is NOT prime"
....:     print "Number of primes: " + str(P)
....:     print "Percentage of primes: " + str(float((P * 100) / (end - start + 1)))
\end{Verbatim}

Here, we compute the values of $f(n) = n^2 - 9n + 61$ for $n = 0, 1,
2, \dots, 50$ and verify the primality of the generated integers:

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: quadratic_prime_formula(0, 50)
 N -- N^2 - 9*N + 61
0 -- 61 is prime
1 -- 53 is prime
2 -- 47 is prime
3 -- 43 is prime
4 -- 41 is prime
5 -- 41 is prime
6 -- 43 is prime
7 -- 47 is prime
8 -- 53 is prime
9 -- 61 is prime
10 -- 71 is prime
11 -- 83 is prime
12 -- 97 is prime
13 -- 113 is prime
14 -- 131 is prime
15 -- 151 is prime
16 -- 173 is prime
17 -- 197 is prime
18 -- 223 is prime
19 -- 251 is prime
20 -- 281 is prime
21 -- 313 is prime
22 -- 347 is prime
23 -- 383 is prime
24 -- 421 is prime
25 -- 461 is prime
26 -- 503 is prime
27 -- 547 is prime
28 -- 593 is prime
29 -- 641 is prime
30 -- 691 is prime
31 -- 743 is prime
32 -- 797 is prime
33 -- 853 is prime
34 -- 911 is prime
35 -- 971 is prime
36 -- 1033 is prime
37 -- 1097 is prime
38 -- 1163 is prime
39 -- 1231 is prime
40 -- 1301 is prime
41 -- 1373 is prime
42 -- 1447 is prime
43 -- 1523 is prime
44 -- 1601 is prime
45 -- 1681 is NOT prime
46 -- 1763 is NOT prime
47 -- 1847 is prime
48 -- 1933 is prime
49 -- 2021 is NOT prime
50 -- 2111 is prime
Number of primes: 48
Percentage of primes: 94.1176470588
\end{Verbatim}

\noindent The last two lines of the output show that $f(n)$ generates
48 primes when $0 \leq n \leq 50$, which is approximately 94\% of the
values generated by $f(n)$ for $n = 0, 2, 3, \dots, 50$. For larger
sequences, it is impractical to print messages indicating the
primality of integers. In the following Sage session, we count the
number of primes generated by $f(n)$ where $0 \leq n \leq 1000$ and
suppress primality messages.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: quadratic_prime_formula(0, 1000, verbose=False)
N -- N^2 - 9*N + 61
Number of primes: 584
Percentage of primes: 58.3416583417
\end{Verbatim}
